---
title: "anchovy report: `r params$sample`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
params:
  # Point this at a run by giving the directory the outputs live in and the
  # sample prefix (the part before _annot_v3.csv etc.). Everything else is
  # derived. Override at render time, e.g.:
  #   rmarkdown::render("anchovy_report.Rmd",
  #     params = list(results_dir = "results/EVA71", sample = "5_EVA71_6h_P5.ccs"))
  results_dir: "."
  sample: "5_EVA71_6h_P5.ccs"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 4.5, dpi = 110)

library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(scales)

theme_set(theme_minimal(base_size = 12) +
          theme(panel.grid.minor = element_blank()))

# --- Locate the output files -------------------------------------------------
# Every panel guards on the file it needs, so a run that skipped a stage (no
# GFF, no pileup counts) still renders -- the dependent sections just say so
# rather than erroring.
P <- function(suffix) file.path(params$results_dir,
                                paste0(params$sample, suffix))

paths <- list(
  annot   = P("_annot_v3.csv"),
  regions = P("_regionAnnotations.csv"),
  nodes   = P("_genotypeNodes.csv"),
  epi     = P("_epistaticNetwork.csv"),
  geno    = P("_genotypeNetwork.csv"),
  allele  = P("_alleleFrequencies.csv"),
  cell    = P("_cellAlleleFreq.csv")
)
have <- function(key) file.exists(paths[[key]])

# read_csv leaves the frequency columns numeric already, but some are written as
# very long decimals; the pipeline also writes some numeric-looking fields that
# arrive as character when a column is all-empty (e.g. `region` with no GFF).
# Coerce defensively at the point of use rather than trusting the header sniff.
num <- function(x) suppressWarnings(as.numeric(x))
```

```{r load}
annot   <- if (have("annot"))   read_csv(paths$annot,   show_col_types = FALSE) else NULL
regions <- if (have("regions")) read_csv(paths$regions, show_col_types = FALSE) else NULL
nodes   <- if (have("nodes"))   read_csv(paths$nodes,   show_col_types = FALSE) else NULL
epi     <- if (have("epi"))     read_csv(paths$epi,     show_col_types = FALSE) else NULL
allele  <- if (have("allele"))  read_csv(paths$allele,  show_col_types = FALSE) else NULL
cellfrq <- if (have("cell"))    read_csv(paths$cell,    show_col_types = FALSE) else NULL

# DEPTH -- the number of cells the whole analysis rests on. `total` in the annot
# table is that count (it is constant down the column).
depth <- if (!is.null(annot)) suppressWarnings(max(annot$total, na.rm = TRUE)) else NA
```

This report summarizes one `anchovy` run: sample **`r params$sample`**, from
`r params$results_dir`. It has two parts. **Quality control** describes the
cell population the results rest on -- how many cells, how well covered, how the
genotype subset compares to the whole. **Results** describes what was found --
where the mutations are, what they do to the protein, and how the genotypes
relate.

```{r overview}
files_present <- vapply(names(paths), have, logical(1))
tibble::tibble(output = names(paths),
               file = basename(unlist(paths)),
               present = ifelse(files_present, "yes", "— not in this run")) |>
  knitr::kable(caption = "Outputs found for this run")
```

# Quality control

## How many cells, and how many carried variants

```{r qc-cellcount, fig.height=3}
if (is.null(annot)) {
  cat("_No annotation table; cannot summarize the cell population._")
} else {
  per_cell <- annot |>
    distinct(CBC_ID, genotype) |>
    mutate(kind = ifelse(genotype == "reference" | is.na(genotype) | genotype == "",
                         "reference (no called variants)", "carries variants"))
  per_cell |>
    count(kind) |>
    ggplot(aes(n, kind, fill = kind)) +
    geom_col(width = 0.6, show.legend = FALSE) +
    geom_text(aes(label = comma(n)), hjust = -0.1) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(x = "cells", y = NULL,
         title = sprintf("%s cells passed filtering", comma(depth)))
}
```

`r if (!is.null(annot)) sprintf("Of %s cells, %s carried at least one called variant.", comma(depth), comma(sum(annot |> distinct(CBC_ID, genotype) |> pull(genotype) |> (\\(g) g != "reference" & !is.na(g) & g != "")())))`

## Mutation load per cell

How many mutations each cell carries. A long right tail can signal artifacts
(the annotate stage drops cells above `max_mutations_per_cell`); a spike at zero
is the reference population above.

```{r qc-mutload, fig.height=3.2}
if (is.null(annot)) {
  cat("_No annotation table._")
} else {
  load_per_cell <- annot |>
    distinct(CBC_ID, genotype) |>
    mutate(n_mut = ifelse(genotype == "reference" | is.na(genotype) | genotype == "",
                          0L, lengths(strsplit(genotype, "_"))))
  ggplot(load_per_cell, aes(n_mut)) +
    geom_bar(fill = "steelblue") +
    scale_x_continuous(breaks = pretty_breaks()) +
    labs(x = "called mutations in the cell", y = "cells",
         title = "Mutation load per cell")
}
```

## Is the genotype subset representative?

The genotype network is built from a strict, well-covered subset of cells; the
allele-frequency table reports the whole population. If the subset is biased,
its per-allele frequencies will pull away from the full population's. Points on
the diagonal mean the subset represents the whole; systematic departure is worth
knowing before trusting the network as a population summary.

```{r qc-subset, fig.height=4.5}
if (is.null(allele) || !all(c("freq_cells","freq_cells_subset") %in% names(allele))) {
  cat("_No allele-frequency table with subset columns; skipping._")
} else {
  d <- allele |>
    transmute(full = num(freq_cells), subset = num(freq_cells_subset)) |>
    filter(is.finite(full), is.finite(subset))
  if (nrow(d) == 0) {
    cat("_No overlapping alleles between subset and full population._")
  } else {
    ggplot(d, aes(full, subset)) +
      geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey60") +
      geom_point(alpha = 0.4, colour = "steelblue") +
      coord_equal() +
      labs(x = "frequency in all cells", y = "frequency in the genotype subset",
           title = "Genotype subset vs. full population, per allele")
  }
}
```

# Results

## Where the mutations are

Every called allele laid out along the genome, colored by whether it changes the
protein. Height is the fraction of cells carrying it (one cell, one vote). A tall
point is a mutation that swept much of the population; the carpet of low points
is the standing diversity. Region bands come from the GFF, if one was used.

```{r results-genomemap, fig.height=4.5}
if (is.null(allele)) {
  cat("_No allele-frequency table; skipping the genome map._")
} else {
  d <- allele |>
    mutate(position = num(position), freq_cells = num(freq_cells)) |>
    filter(is.finite(position), is.finite(freq_cells))

  # Classify syn/non-syn from the region annotation if present; the allele table
  # itself does not carry sub_class, so join it in by genome position + allele.
  if (!is.null(regions)) {
    cls <- regions |>
      filter(region_type == "coding") |>
      transmute(position = genome_pos, allele = mut_base,
                sub_class = sub_class) |>
      distinct(position, allele, .keep_all = TRUE)
    d <- d |> left_join(cls, by = c("position", "allele")) |>
      mutate(sub_class = ifelse(is.na(sub_class), "non-coding / unannotated", sub_class))
  } else {
    d$sub_class <- "unclassified (no GFF)"
  }

  pal <- c("Non-Syn" = "#d1495b", "Syn" = "#66a182",
           "non-coding / unannotated" = "grey55",
           "unclassified (no GFF)" = "grey55", "X" = "grey80")
  ggplot(d, aes(position, freq_cells, colour = sub_class)) +
    geom_point(alpha = 0.7, size = 1.6) +
    scale_colour_manual(values = pal, name = NULL) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(x = "genome position (nt)", y = "fraction of cells",
         title = "Allele frequency across the genome") +
    theme(legend.position = "top")
}
```

## Synonymous vs. non-synonymous

The balance of protein-changing to silent mutations, overall and per region.
A region skewed toward non-synonymous changes is under different selective
pressure than one that is mostly silent.

```{r results-dnds, fig.height=4}
if (is.null(regions)) {
  cat("_No region annotations (no GFF used); skipping syn/non-syn breakdown._")
} else {
  d <- regions |>
    filter(region_type == "coding", sub_class %in% c("Syn", "Non-Syn")) |>
    # count each distinct genomic mutation once per region it falls in
    distinct(region, genome_pos, mut_base, sub_class)
  if (nrow(d) == 0) {
    cat("_No coding syn/non-syn calls to summarize._")
  } else {
    d |>
      count(region, sub_class) |>
      ggplot(aes(n, reorder(region, n, sum), fill = sub_class)) +
      geom_col() +
      scale_fill_manual(values = c("Non-Syn" = "#d1495b", "Syn" = "#66a182"),
                        name = NULL) +
      labs(x = "distinct mutations", y = NULL,
           title = "Coding mutations by region") +
      theme(legend.position = "top")
  }
}
```

## Genotype network

Each node is a genotype; size is how many cells carry it, colour is how many
mutations it carries. Edges connect genotypes differing by a single mutation
(the epistatic network) -- the paths the population actually took through
sequence space. Only the connected part is drawn; singletons with no single-step
neighbour are omitted for legibility (they remain in `genotypeNodes.csv`).

```{r results-network, fig.height=6.5, fig.width=9}
have_graph <- requireNamespace("igraph", quietly = TRUE) &&
              requireNamespace("ggraph", quietly = TRUE)

if (is.null(epi) || is.null(nodes)) {
  cat("_Network or node table missing; skipping._")
} else if (!have_graph) {
  cat("_Packages `igraph` and `ggraph` are not installed. Add them with:_\n\n",
      "```\nconda install -c conda-forge r-igraph r-ggraph\n```\n\n",
      "_Meanwhile the network CSVs open directly in Cytoscape (see the README)._")
} else {
  library(igraph); library(ggraph)

  edges <- epi |>
    filter(source != target, target != "reference") |>
    distinct(source, target)

  if (nrow(edges) == 0) {
    cat("_No genotype-to-genotype single-step edges in this run._")
  } else {
    g <- graph_from_data_frame(edges, directed = FALSE,
                               vertices = NULL)
    # attach node attributes where we have them
    nd <- nodes |> filter(genotype %in% V(g)$name)
    idx <- match(V(g)$name, nd$genotype)
    V(g)$nCells     <- nd$nCells[idx]
    V(g)$nMutations <- nd$nMutations[idx]
    V(g)$nCells[is.na(V(g)$nCells)] <- 1
    V(g)$nMutations[is.na(V(g)$nMutations)] <- 0

    ggraph(g, layout = "fr") +
      geom_edge_link(alpha = 0.25) +
      geom_node_point(aes(size = nCells, colour = nMutations)) +
      scale_size_continuous(range = c(1.5, 8), name = "cells") +
      scale_colour_viridis_c(name = "mutations") +
      labs(title = "Genotype network (single-step edges)") +
      theme_void(base_size = 12) +
      theme(legend.position = "right")
  }
}
```

## The most common genotypes

The genotypes carried by the most cells, with their amino-acid-level names.
A mutation appearing in nearly every genotype is one that had already swept the
population when it was sampled.

```{r results-topgeno}
if (is.null(nodes)) {
  cat("_No node table._")
} else {
  nodes |>
    mutate(nCells = num(nCells), genoFreq = num(genoFreq)) |>
    arrange(desc(nCells)) |>
    slice_head(n = 15) |>
    transmute(genotype = genotypeName, mutations = nMutations,
              cells = nCells, `freq` = percent(genoFreq, accuracy = 0.1)) |>
    knitr::kable(caption = "Top 15 genotypes by cell count")
}
```

## Within-cell allele frequencies

Where the pileup counts were produced (`sam2consensus --counts`), this shows
alleles present at intermediate frequency *within* individual cells -- candidate
mixed infections or within-cell diversity, as opposed to differences between
cells. Most calls sit near 1.0 (the cell's consensus base); a spread below that
is the within-cell signal.

```{r results-withincell, fig.height=3.5}
if (is.null(cellfrq)) {
  cat("_No per-cell allele table (run `anchovy frequencies --counts-dir` to produce it)._")
} else {
  cellfrq |>
    mutate(freq = num(freq)) |>
    filter(is.finite(freq)) |>
    ggplot(aes(freq)) +
    geom_histogram(bins = 40, fill = "steelblue") +
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    labs(x = "within-cell allele frequency", y = "allele observations",
         title = "Within-cell allele frequency spectrum")
}
```

---

```{r sessioninfo}
cat("Rendered", format(Sys.time(), "%Y-%m-%d %H:%M"),
    "\nanchovy report for", params$sample)
```
