---
title: "anchovy report: `r params$sample`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
params:
  # Point this at a run by giving the directory the outputs live in and the
  # sample prefix (the part before _annot_v3.csv etc.). Everything else is
  # derived. Override at render time, e.g.:
  #   rmarkdown::render("anchovy_report.Rmd",
  #     params = list(results_dir = "results/EVA71", sample = "5_EVA71_6h_P5.ccs"))
  results_dir: "."
  sample: "5_EVA71_6h_P5.ccs"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 4.5, dpi = 110)

library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(scales)

theme_set(theme_minimal(base_size = 12) +
          theme(panel.grid.minor = element_blank()))

# --- Locate the output files -------------------------------------------------
# Every panel guards on the file it needs, so a run that skipped a stage (no
# GFF, no pileup counts) still renders -- the dependent sections just say so
# rather than erroring.
P <- function(suffix) file.path(params$results_dir,
                                paste0(params$sample, suffix))

paths <- list(
  annot   = P("_annot_v3.csv"),
  regions = P("_regionAnnotations.csv"),
  nodes   = P("_genotypeNodes.csv"),
  epi     = P("_epistaticNetwork.csv"),
  geno    = P("_genotypeNetwork.csv"),
  allele  = P("_alleleFrequencies.csv"),
  cell    = P("_cellAlleleFreq.csv"),
  cons    = P("_allConsensus.fasta")
)
have <- function(key) file.exists(paths[[key]])

# read_csv leaves the frequency columns numeric already, but some are written as
# very long decimals; the pipeline also writes some numeric-looking fields that
# arrive as character when a column is all-empty (e.g. `region` with no GFF).
# Coerce defensively at the point of use rather than trusting the header sniff.
num <- function(x) suppressWarnings(as.numeric(x))

# --- Substitution-class colours, used by every panel that splits on it --------
# Blue/orange rather than the red/green this report used to carry. That pair
# put a deuteranope -- the commonest colour-vision deficiency, ~6% of men --
# at a CVD separation of 6.5 (OKLab x100, 8 is the target), on the one
# distinction those panels exist to draw; the green also sat under the chroma
# floor, reading grey, and under 3:1 contrast on white. This pair measures 24.7
# CVD and 33.6 normal-vision separation.
#
# Defined once so the raster, the genome scatter and the region bars cannot
# drift apart: the colour follows the substitution class, not the panel.
SUBCLASS_PAL <- c(
  "Non-Syn"                  = "#2a78d6",
  "Syn"                      = "#eb6834",
  # Neutrals, deliberately: these mean "no call", not another category.
  "non-coding / unannotated" = "grey55",
  "unclassified (no GFF)"    = "grey55",
  "X"                        = "grey80"
)
```

```{r load}
annot   <- if (have("annot"))   read_csv(paths$annot,   show_col_types = FALSE) else NULL
regions <- if (have("regions")) read_csv(paths$regions, show_col_types = FALSE) else NULL
nodes   <- if (have("nodes"))   read_csv(paths$nodes,   show_col_types = FALSE) else NULL
epi     <- if (have("epi"))     read_csv(paths$epi,     show_col_types = FALSE) else NULL
allele  <- if (have("allele"))  read_csv(paths$allele,  show_col_types = FALSE) else NULL
cellfrq <- if (have("cell"))    read_csv(paths$cell,    show_col_types = FALSE) else NULL

# DEPTH -- the number of cells the whole analysis rests on. `total` in the annot
# table is that count (it is constant down the column).
depth <- if (!is.null(annot)) suppressWarnings(max(annot$total, na.rm = TRUE)) else NA

# --- Per-cell sequencing depth, from the consensus FASTA headers -------------
# sam2consensus writes "coverage:NN length:MM" onto every consensus it emits,
# and that is the ONLY per-cell depth record surviving the pipeline -- no CSV
# downstream carries it. Parsed here rather than recomputed, since the reads it
# came from are long gone by this stage.
#
# These cover every cell that produced a consensus, not just the filtered
# subset the annotation table describes, so this section sees the population
# the filters were applied TO.
parse_consensus_headers <- function(path) {
  ln <- readLines(path, warn = FALSE)
  hi <- which(startsWith(ln, ">"))
  if (!length(hi)) return(NULL)
  h <- ln[hi]
  tibble::tibble(
    cell     = sub("^>([^ ]+).*", "\\1", h),
    coverage = num(sub(".*coverage:([0-9.]+).*", "\\1", h)),
    called   = num(sub(".*length:([0-9]+).*",   "\\1", h))
  )
}

# Reference length, taken from the consensuses themselves so it needs no extra
# input. Sums each record's block, in case the FASTA is wrapped.
#
# THE MODAL LENGTH, NOT THE FIRST RECORD'S. sam2consensus emits a called
# insertion as an extra column, so a record carrying one is LONGER than the
# reference -- and if that record happens to be first, using it as the
# denominator understates every cell's breadth. On this dataset about 4% of
# cells were insertion-bearing, so it is a live possibility rather than a
# theoretical one. The population's most common length is the reference.
consensus_ref_length <- function(path) {
  ln <- readLines(path, warn = FALSE)
  hi <- which(startsWith(ln, ">"))
  if (!length(hi)) return(NA_real_)
  ends <- c(hi[-1] - 1L, length(ln))
  lens <- vapply(seq_along(hi), function(i) {
    if (ends[i] <= hi[i]) return(0)
    as.numeric(sum(nchar(ln[(hi[i] + 1L):ends[i]])))
  }, numeric(1))
  lens <- lens[lens > 0]
  if (!length(lens)) return(NA_real_)
  as.numeric(names(sort(table(lens), decreasing = TRUE))[1])
}

covstats   <- if (have("cons")) parse_consensus_headers(paths$cons) else NULL
genome_len <- if (have("cons")) consensus_ref_length(paths$cons)    else NA_real_
```

This report summarizes one `anchovy` run: sample **`r params$sample`**, from
`r params$results_dir`. It has two parts. **Quality control** describes the
cell population the results rest on -- how many cells, how well covered, how the
genotype subset compares to the whole. **Results** describes what was found --
where the mutations are, what they do to the protein, and how the genotypes
relate.

```{r overview}
files_present <- vapply(names(paths), have, logical(1))
tibble::tibble(output = names(paths),
               file = basename(unlist(paths)),
               present = ifelse(files_present, "yes", "— not in this run")) |>
  knitr::kable(caption = "Outputs found for this run")
```

# Quality control

## Coverage and depth

What the calls rest on. These cover every cell that produced a consensus --
including the ones later filtered out -- so this is the population the depth
and completeness filters were applied to, not what survived them.

**`coverage` is not a depth.** sam2consensus computes it as summed depth over
positions with reads, divided by the FULL reference length, so it falls when a
cell spans less of the genome however deep those reads were. A cell at 40x over
40% of the genome scores 16. The two panels below therefore separate it: mean
coverage as written, and how much of the genome the cell actually called.

```{r qc-coverage, fig.height=3.2}
if (is.null(covstats) || !nrow(covstats)) {
  cat("_No `_allConsensus.fasta`; skipping per-cell coverage._")
} else {
  med <- median(covstats$coverage, na.rm = TRUE)
  ggplot(covstats, aes(coverage)) +
    geom_histogram(bins = 50, fill = "steelblue") +
    geom_vline(xintercept = med, linetype = "dashed", colour = "grey30") +
    annotate("text", x = med, y = Inf, vjust = 1.4, hjust = -0.1, size = 3.4,
             colour = "grey30", label = sprintf("median %.1f", med)) +
    labs(x = "mean coverage over the whole reference", y = "cells",
         title = sprintf("Per-cell coverage (%s cells with a consensus)",
                         comma(nrow(covstats))))
}
```

```{r qc-breadth, fig.height=3.2}
if (is.null(covstats) || !nrow(covstats) || is.na(genome_len)) {
  cat("_No `_allConsensus.fasta`; skipping per-cell breadth._")
} else {
  b <- covstats |> mutate(breadth = called / genome_len)
  med <- median(b$breadth, na.rm = TRUE)
  ggplot(b, aes(breadth)) +
    geom_histogram(bins = 50, fill = "darkseagreen4") +
    geom_vline(xintercept = med, linetype = "dashed", colour = "grey30") +
    # No hard 0-1 limit: an insertion-bearing record can call more positions
    # than the modal reference length, and ggplot drops out-of-limit points
    # silently -- losing cells from the panel that exists to count them.
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    labs(x = "fraction of the reference the cell called", y = "cells",
         title = sprintf("Per-cell breadth (reference %s nt, median %.0f%%)",
                         comma(genome_len), 100 * med))
}
```

`r if (!is.null(covstats) && nrow(covstats) && !is.na(genome_len)) sprintf("Depth where a base was actually called, rather than averaged over the whole reference, has median %.1fx -- the number the coverage panel understates for partially covered cells.", median(covstats$coverage * genome_len / pmax(covstats$called, 1), na.rm = TRUE))`

### Along the genome

Per-position read depth and how many cells had a confident call there. Sampled
at the positions carrying a called allele, so the x-axis is not every base --
but it is dense enough to show where coverage falls away, which is usually the
ragged 5'/3' flanks every amplicon run has.

```{r qc-depth-profile, fig.height=3.6}
if (is.null(allele) || !all(c("position", "cells_called") %in% names(allele))) {
  cat("_No allele-frequency table; skipping the coverage profile._")
} else {
  # reads_depth is only populated when the run dumped pileup counts, and an
  # older output may not carry the column at all -- so build from what is
  # there rather than naming it unconditionally.
  prof <- allele |>
    distinct(position, .keep_all = TRUE) |>
    transmute(position = num(position),
              `cells with a call` = num(cells_called))
  if ("reads_depth" %in% names(allele)) {
    prof$`read depth` <- num(allele |> distinct(position, .keep_all = TRUE) |>
                             pull(reads_depth))
  }
  prof <- prof |>
    pivot_longer(-position, names_to = "measure", values_to = "value") |>
    filter(!is.na(value))
  if (!nrow(prof)) {
    cat("_No per-position depth recorded._")
  } else {
    ggplot(prof, aes(position, value)) +
      geom_line(linewidth = 0.4, colour = "grey25") +
      facet_wrap(~measure, ncol = 1, scales = "free_y") +
      scale_y_continuous(labels = comma) +
      labs(x = "genome position (nt)", y = NULL,
           title = "Coverage along the genome")
  }
}
```

## How many cells, and how many carried variants

```{r qc-cellcount, fig.height=3}
if (is.null(annot)) {
  cat("_No annotation table; cannot summarize the cell population._")
} else {
  per_cell <- annot |>
    distinct(CBC_ID, genotype) |>
    mutate(kind = ifelse(genotype == "reference" | is.na(genotype) | genotype == "",
                         "reference (no called variants)", "carries variants"))
  per_cell |>
    count(kind) |>
    ggplot(aes(n, kind, fill = kind)) +
    geom_col(width = 0.6, show.legend = FALSE) +
    geom_text(aes(label = comma(n)), hjust = -0.1) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(x = "cells", y = NULL,
         title = sprintf("%s cells passed filtering", comma(depth)))
}
```

`r if (!is.null(annot)) sprintf("Of %s cells, %s carried at least one called variant.", comma(depth), comma(sum(annot |> distinct(CBC_ID, genotype) |> pull(genotype) |> (\(g) g != "reference" & !is.na(g) & g != "")())))`

## Mutation load per cell

How many mutations each cell carries. A long right tail can signal artifacts
(the annotate stage drops cells above `max_mutations_per_cell`); a spike at zero
is the reference population above.

```{r qc-mutload, fig.height=3.2}
if (is.null(annot)) {
  cat("_No annotation table._")
} else {
  load_per_cell <- annot |>
    distinct(CBC_ID, genotype) |>
    mutate(n_mut = ifelse(genotype == "reference" | is.na(genotype) | genotype == "",
                          0L, lengths(strsplit(genotype, "_"))))
  ggplot(load_per_cell, aes(n_mut)) +
    geom_bar(fill = "steelblue") +
    scale_x_continuous(breaks = pretty_breaks()) +
    labs(x = "called mutations in the cell", y = "cells",
         title = "Mutation load per cell")
}
```

## Is the genotype subset representative?

The genotype network is built from a strict, well-covered subset of cells; the
allele-frequency table reports the whole population. If the subset is biased,
its per-allele frequencies will pull away from the full population's. Points on
the diagonal mean the subset represents the whole; systematic departure is worth
knowing before trusting the network as a population summary.

```{r qc-subset, fig.height=4.5}
if (is.null(allele) || !all(c("freq_cells","freq_cells_subset") %in% names(allele))) {
  cat("_No allele-frequency table with subset columns; skipping._")
} else {
  d <- allele |>
    transmute(full = num(freq_cells), subset = num(freq_cells_subset)) |>
    filter(is.finite(full), is.finite(subset))
  if (nrow(d) == 0) {
    cat("_No overlapping alleles between subset and full population._")
  } else {
    ggplot(d, aes(full, subset)) +
      geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey60") +
      geom_point(alpha = 0.4, colour = "steelblue") +
      coord_equal() +
      labs(x = "frequency in all cells", y = "frequency in the genotype subset",
           title = "Genotype subset vs. full population, per allele")
  }
}
```

# Results

## Where the mutations are

Every called allele laid out along the genome, colored by whether it changes the
protein. Height is the fraction of cells carrying it (one cell, one vote). A tall
point is a mutation that swept much of the population; the carpet of low points
is the standing diversity. Region bands come from the GFF, if one was used.

```{r results-genomemap, fig.height=4.5}
if (is.null(allele)) {
  cat("_No allele-frequency table; skipping the genome map._")
} else {
  d <- allele |>
    mutate(position = num(position), freq_cells = num(freq_cells)) |>
    filter(is.finite(position), is.finite(freq_cells))

  # Classify syn/non-syn from the region annotation if present; the allele table
  # itself does not carry sub_class, so join it in by genome position + allele.
  if (!is.null(regions)) {
    cls <- regions |>
      filter(region_type == "coding") |>
      transmute(position = genome_pos, allele = mut_base,
                sub_class = sub_class) |>
      distinct(position, allele, .keep_all = TRUE)
    d <- d |> left_join(cls, by = c("position", "allele")) |>
      mutate(sub_class = ifelse(is.na(sub_class), "non-coding / unannotated", sub_class))
  } else {
    d$sub_class <- "unclassified (no GFF)"
  }

  ggplot(d, aes(position, freq_cells, colour = sub_class)) +
    geom_point(alpha = 0.7, size = 1.6) +
    scale_colour_manual(values = SUBCLASS_PAL, name = NULL) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(x = "genome position (nt)", y = "fraction of cells",
         title = "Allele frequency across the genome") +
    theme(legend.position = "top")
}
```

## Every cell, every mutation

The same genome axis as above, resolved per cell: one row per cell, one point
per mutation in that cell's consensus. Cells are ordered by how many mutations
they carry and then by genotype, so cells sharing a genotype sit together and a
mutation the population shares reads as a vertical line down the plot.

A blank row is a cell whose consensus matched the reference — those are kept,
because how much of the plot is empty is part of what it shows.

```{r results-cellmap, fig.height=8}
if (is.null(annot)) {
  cat("_No annotation table; skipping the per-cell map._")
} else {
  # EVERY cell gets a row, including the ones carrying nothing. They are the
  # rows with no points, and dropping them would make the population look more
  # mutated than it is. They arrive here as the annot table's reference rows,
  # which have no position -- hence building the row order from the cell list
  # and the points from a separate, filtered frame.
  per_cell <- annot |>
    distinct(CBC_ID, genotype) |>
    mutate(n_mut = ifelse(genotype == "reference" | is.na(genotype) |
                            genotype == "",
                          0L, lengths(strsplit(genotype, "_")))) |>
    arrange(n_mut, genotype, CBC_ID)

  pts <- annot |>
    filter(!is.na(pos)) |>
    transmute(CBC_ID,
              pos = num(pos),
              # Anything the annotator could not classify is drawn in the
              # neutral, not left unmapped -- scale_colour_manual gives an
              # unmapped level no colour, which would silently drop the point.
              sub_class = ifelse(subClass %in% names(SUBCLASS_PAL),
                                 subClass, "X"))

  # drop = FALSE plus the full level set is what keeps the empty rows: the
  # points frame only contains cells that have mutations.
  pts$CBC_ID <- factor(pts$CBC_ID, levels = per_cell$CBC_ID)

  ggplot(pts, aes(pos, CBC_ID, colour = sub_class)) +
    geom_point(shape = 15, size = 0.45) +
    scale_colour_manual(values = SUBCLASS_PAL, name = NULL) +
    scale_y_discrete(drop = FALSE) +
    labs(x = "genome position (nt)",
         y = sprintf("%s cells, fewest mutations at the bottom",
                     comma(nrow(per_cell))),
         title = "Mutations per cell across the genome") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major.y = element_blank(),
          panel.grid.minor = element_blank(),
          legend.position = "top")
}
```

`r if (!is.null(annot)) sprintf("At %s cells in one panel each row is under two pixels tall, so read the vertical structure rather than individual rows -- a column of points is a mutation many cells share, and a block of identical rows is one genotype recurring.", comma(annot |> distinct(CBC_ID) |> nrow()))`

## Synonymous vs. non-synonymous

The balance of protein-changing to silent mutations, overall and per region.
A region skewed toward non-synonymous changes is under different selective
pressure than one that is mostly silent.

```{r results-dnds, fig.height=4}
if (is.null(regions)) {
  cat("_No region annotations (no GFF used); skipping syn/non-syn breakdown._")
} else {
  d <- regions |>
    filter(region_type == "coding", sub_class %in% c("Syn", "Non-Syn")) |>
    # count each distinct genomic mutation once per region it falls in
    distinct(region, genome_pos, mut_base, sub_class)
  if (nrow(d) == 0) {
    cat("_No coding syn/non-syn calls to summarize._")
  } else {
    d |>
      count(region, sub_class) |>
      ggplot(aes(n, reorder(region, n, sum), fill = sub_class)) +
      geom_col() +
      scale_fill_manual(values = SUBCLASS_PAL, name = NULL) +
      labs(x = "distinct mutations", y = NULL,
           title = "Coding mutations by region") +
      theme(legend.position = "top")
  }
}
```

## Genotype network

Each node is a genotype; size is how many cells carry it, colour is how many
mutations it carries. Edges connect genotypes differing by a single mutation
(the epistatic network) -- the paths the population actually took through
sequence space. Only the connected part is drawn; singletons with no single-step
neighbour are omitted for legibility (they remain in `genotypeNodes.csv`).

```{r results-network, fig.height=6.5, fig.width=9}
have_graph <- requireNamespace("igraph", quietly = TRUE) &&
              requireNamespace("ggraph", quietly = TRUE)

if (is.null(epi) || is.null(nodes)) {
  cat("_Network or node table missing; skipping._")
} else if (!have_graph) {
  cat("_Packages `igraph` and `ggraph` are not installed. Add them with:_\n\n",
      "```\nconda install -c conda-forge r-igraph r-ggraph\n```\n\n",
      "_Meanwhile the network CSVs open directly in Cytoscape (see the README)._")
} else {
  library(igraph); library(ggraph)

  edges <- epi |>
    filter(source != target, target != "reference") |>
    distinct(source, target)

  if (nrow(edges) == 0) {
    cat("_No genotype-to-genotype single-step edges in this run._")
  } else {
    g <- graph_from_data_frame(edges, directed = FALSE,
                               vertices = NULL)
    # attach node attributes where we have them
    nd <- nodes |> filter(genotype %in% V(g)$name)
    idx <- match(V(g)$name, nd$genotype)
    V(g)$nCells     <- nd$nCells[idx]
    V(g)$nMutations <- nd$nMutations[idx]
    V(g)$nCells[is.na(V(g)$nCells)] <- 1
    V(g)$nMutations[is.na(V(g)$nMutations)] <- 0

    ggraph(g, layout = "fr") +
      geom_edge_link(alpha = 0.25) +
      geom_node_point(aes(size = nCells, colour = nMutations)) +
      scale_size_continuous(range = c(1.5, 8), name = "cells") +
      scale_colour_viridis_c(name = "mutations") +
      labs(title = "Genotype network (single-step edges)") +
      theme_void(base_size = 12) +
      theme(legend.position = "right")
  }
}
```

## The most common genotypes

The genotypes carried by the most cells, with their amino-acid-level names.
A mutation appearing in nearly every genotype is one that had already swept the
population when it was sampled.

```{r results-topgeno}
if (is.null(nodes)) {
  cat("_No node table._")
} else {
  nodes |>
    mutate(nCells = num(nCells), genoFreq = num(genoFreq)) |>
    arrange(desc(nCells)) |>
    slice_head(n = 15) |>
    transmute(genotype = genotypeName, mutations = nMutations,
              cells = nCells, `freq` = percent(genoFreq, accuracy = 0.1)) |>
    knitr::kable(caption = "Top 15 genotypes by cell count")
}
```

## Within-cell allele frequencies

Where the pileup counts were produced (`sam2consensus --counts`), this shows
alleles present at intermediate frequency *within* individual cells -- candidate
mixed infections or within-cell diversity, as opposed to differences between
cells. Most calls sit near 1.0 (the cell's consensus base); a spread below that
is the within-cell signal.

```{r results-withincell, fig.height=3.5}
if (is.null(cellfrq)) {
  cat("_No per-cell allele table (run `anchovy frequencies --counts-dir` to produce it)._")
} else {
  cellfrq |>
    mutate(freq = num(freq)) |>
    filter(is.finite(freq)) |>
    ggplot(aes(freq)) +
    geom_histogram(bins = 40, fill = "steelblue") +
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    labs(x = "within-cell allele frequency", y = "allele observations",
         title = "Within-cell allele frequency spectrum")
}
```

---

```{r sessioninfo}
cat("Rendered", format(Sys.time(), "%Y-%m-%d %H:%M"),
    "\nanchovy report for", params$sample)
```
